Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-06-18 report/emse24.tex
feat(report): emse24 updates

  \todo{update the numbers here}
Testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part. Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML. And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.

@@

Machine Learning Development Lifecycle (MLDL) encompasses multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves Exploratory Data Analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analyzing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\cite{haakman2021ai,amershi2019software,sculley2015hidden}.

@@

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles which emphasize incorporating early feedback and working in small, manageable iterations. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects~\cite{betz2018managing,martinez-plumed2021crisp-dm}.

Computational notebooks---specifically Jupyter notebooks~\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}. Jupyter notebooks provide an interactive computing environment that allow users to break complex code into smaller, more manageable chunks. Each notebook is composed of a series of code cells, which can be executed independently in a Read-Eval-Print Loop (REPL) style of development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\cite{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.

Feedback is crucial for making important decisions throughout the MLDL. Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, create new features to boost model performance, and select specific hyperparameters. However, these decisions are often based on the conditions present when the model was initially created. Visual outputs in particular, are open to interpretation~\cite{heer2010tour}. The viewer's background and expertise significantly influence how they understand and draw conclusions from visual data. 
For instance, interpreting a histogram requires knowledge of various distribution types, which someone without statistical training might not possess. Additionally, physical limitations such as color blindness can impede accurate interpretation if visualizations are not designed with accessibility in mind. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

@@

To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct an empirical study of the outputs of code cells in Jupyter notebooks. Next, we conduct an empirical study of Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks. Finally, we investigate the intersection of outputs and assertions by conducting an empirical analysis of assertions that document decisions made from the outputs of code cells. In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt.

% TODO: contributions?
% TODO: key observations?

% Visualisations are used throughout the ml development lifecycle to test various properties of the ml system. in the early stages of the ml development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties. during the model development phase, visualisations are used to summarise metrics, contrast different learning algorithms and iteratively fine-tune ml models. once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold~\cite{yuan2021survey,hohman2019visual,amershi2015modeltracker,wexler2020if}.

% our observations indicate that formulating analytical assertions from visualisations is an emerging testing technique. however, the assertions found in this study often do not reflect the information obtained from their visual counter-part. this study finds the need for automated tools that can reduce the manual effort necessary to validate visualisations in ml systems. and help ml practitioners to formulate better analytical assertions from visualisations.

@@

@@

% describe the structure of jupyter notebooks (anatomy: cells, outputs); no need to go into too much technical detail regarding the internal JSON structure, just mention that the data is represented in a standardised structure internally

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{nb.pdf}
  \label{fig:nb}
  \caption{Example Jupyter notebook with code and markdown cells, adapted from~\cite[Figure~1]{pimentel2019large-scale}.}
\end{figure}

Literate programming is a paradigm introduced by Donald Knuth that emphasizes writing code that is understandable by humans. In his seminal work, Knuth adovocates combining source code with documentation in a single, cohesive narrative~\cite{knuth1984literate}. Computational notebooks are inspired by this paradigm, allowing for the integration of executable code, scientific equations, visualizations, and explanatory text within a single document. This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows. Project Jupyter is a prominent example of this approach, offering a suite of open-source tools for interactive computing. Jupyter Notebooks, in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.

% front-end which parses the JSON
% there is also the kernel which is a process running in the background, the frontend communicates with the kernel; send input (code cell) and get output (result of execution)

Figure~\ref{fig:nb} shows an Jupyter notebook with markdown and code cells. The data comprised within Jupyter Notebooks is stored using the JSON schema specified by \emph{nbformat}~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically. A notebook consists of \emph{cells} which can be of three types---``markdown'', ``code'' and ``raw''. All cells of a notebook are stored as a list under the top-level \lstinline[language={}]$cells$ key in the JSON representation. Within this list, each cell is represented as a JSON object. The contents of the cell are stored as a list of strings inside the \lstinline[language={}]$source$ key.

@@

An Abstract Syntax Tree (AST) is a tree representation of the abstract syntactic structure of source code. Each node in the tree represents a construct occurring in the source code. ASTs are widely used in compilers and code analysis tools to understand and manipulate code programmatically~\cite{wikipedia2024abstract}. The Python \emph{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an AST. This module allows for the examination and modification of code structure, making it valuable for tasks such as code analysis, optimization, and transformation. The ast module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation~\cite{pythonast}.

In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class. The \lstinline{Assert} node includes the \lstinline{test} attribute---which contains the condition being asserted---and the \lstinline{msg} attribute---which holds an optional message displayed if the assertion fails. \lstinline{print} statements are treated as a call to the built-in \lstinline{print} function. Function calls are represented in the AST by the \lstinline{Call} class. The \lstinline{Call} node includes \lstinline{func}, representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.

\section{Study Design}

\subsection{Motivating Example}

Sara is a data scientist who is developing a machine learning model to predict customer churn. Sara begins her project in a Jupyter notebook, which supports iterative development and immediate feedback through the execution of code cells. Her workflow is dynamic and exploratory, typical of modern machine learning practices.

Initially, Sara loads a dataset and uses basic descriptive statistics to understand its distribution. As she executes a cell that calculates and displays the mean and standard deviation of each feature, she notices unusually high variance in one of the features. This insight, derived directly from the output of the notebook cell, prompts her to apply normalization to ensure better model performance.

With data preprocessing complete, Sara proceeds to build her initial model. After setting up a logistic regression classifier, she runs a training cell that outputs the model's accuracy and loss after the first epoch. The results are disappointing, prompting her to adjust the learning rate and batch size directly in subsequent cells. Each modification is immediately followed by feedback on its impact, allowing Sara to quickly converge on the optimal settings.

Throughout this process, Sara heavily relies on the outputs of various code cells, not just for error correction but also for making strategic decisions about model architecture, such as adding layers or changing activation functions. At one point, after noticing overfitting from a cell outputting validation loss, she decides to incorporate dropout layers, which she can quickly evaluate by rerunning the training process.

Finally, Sara fine-tunes her model based on the outputs from a confusion matrix and ROC curve, identifying thresholds for classification that balance precision and recall effectively. Each decision in Sara's workflow—from preprocessing adjustments to hyperparameter tuning and model validation—is guided by the iterative feedback she receives from the outputs of her notebook's code cells.

% TODO: the jump to assertions is very drastic; motivate it using scenario that there is a bug in the system (use the data validation papers from google) and Sara needs to fix them and how assertions would have helped here to debug the issue faster! (notebook keeps running!)
Once all experimentation and model tuning are complete, and Sara is ready to deploy the model into a production environment, she adds assertions as a final step. In the dynamic, interactive environment of Jupyter notebooks, where code cells can be executed in any order, there is a risk of arriving at inconsistent states that may affect the model's performance unpredictably. Assertions act as a safeguard against these inconsistencies by verifying that the data fed into the model adheres to expected formats and distributions and that the model's performance metrics meet predetermined standards. For instance, assertions are used to ensure that no data columns have null values, that the range of input data features remains within expected bounds, and that model accuracy thresholds are consistently met regardless of the execution order of the notebook cells. This practice not only reinforces the stability and reliability of the deployed model but also provides clear, executable documentation of the critical decisions made throughout the model's development. When new team members are onboarded, these assertions allow them to quickly grasp the underlying assumptions and operational parameters of the model, ensuring ongoing robustness and performance integrity. This is crucial in maintaining the quality and reliability of the model as it transitions from a development setting to a production environment.

\subsection{Exploratory Analysis}

We undertook an exploratory analysis of the collected data to formulate our research questions.

\subsection{Research Questions}

\begin{description}
  \item[RQ1.] \textbf{How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}
  \item[RQ2.] \textbf{How are assertions used to validate ML code written in Jupyter notebooks?}
  \item[RQ3.] \textbf{How are assertions used to validate visual outputs in Jupyter notebooks?}
\end{description}

@@

% The predominant form of obtaining feedback, while developing ML pipelines in Jupyter notebooks, is through the output of code cells. These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualizations. Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code. This idea draws from prior literature in software engineering that connects test-driven development (TDD) as a form of developing software using feedback.

@@

We check the validity of the underlying JSON structure in each notebook using the nbformat tool ~\footnote{https://nbformat.readthedocs.io/en/latest/}. Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from white literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.

@@

\subsection{(RQ1) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}

@@

Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis. For instance, visualizations can efficiently determine if scaling, normalizing, or handling outliers is needed. During the exploratory data analysis (EDA) phase, visualizations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training. Additionally, descriptive statistics are used post-transformation to verify the effectiveness of steps like data normalization, ensuring data conforms to the expected format for further analysis.

\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualizations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48). These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the EDA phase.}

@@

Table~\ref{tab:linear-relation-check} illustrates two outputs found in this study for identifying linear relationships amongst features in the data. In output \emph{O6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}. Output \emph{O10} depicts the visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.

Linear machine learning models achieve optimal performance when the target variable can be expressed as a linear combination of the input features. However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training. Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\ref{shome2022data}.

\highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data. Visualizations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features. These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}

@@

Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system. This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.

@@

While training multiple machine learning models for a single task is common, choosing the best model considers not only its performance but also its training speed. Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{CITEME}.

@@

Table~\ref{tab:shape-check} shows print statements discovered in this study which check the shape of the data. \emph{P4} prints the number of rows or examples in the test set. \emph{P32} and \emph{117} prints the number of rows and columns (or features) in the training and test data respectively.

@@

\subsection{(RQ2) How are assertions used to validate ML code written in Jupyter notebooks?}~\label{sec:result-rq2}

@@

Table~\ref{tab:assert-resource-check} persents the assertions encountered in this study, that perform resource checks encompassing pre-trained models, data paths, visualizations, and model training status. These checks help prevent runtime errors and ensure the proper execution of code.

@@

Resource check assertions play a vital role in maintaining the reliability and robustness of the machine learning workflows by ensuring the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualizations.

@@

>>>>

  \todo{update the numbers here}
Developing and testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part. Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML. And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.

@@

The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergo cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analyzing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\cite{haakman2021ai,amershi2019software,sculley2015hidden}.

@@

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\cite{betz2018managing}. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, data preparation, modeling, evaluation and finally deployment and monitoring~\cite{martinez-plumed2021crisp-dm}. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.

Computational notebooks---specifically Jupyter notebooks\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}. Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks. Each notebook is composed of a series of code cells, which can be executed independently in a read-eval-print loop (REPL) style of interactive software development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\cite{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.

Feedback is crucial for making important decisions throughout the MLDL. Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and select specific hyperparameters. However, these decisions are often based on the conditions present when the model was initially created. Visual outputs in particular, are open to interpretation~\cite{heer2010tour}. The viewer's background and expertise significantly influence how they understand and draw conclusions from visual data. 
For instance, interpreting a histogram requires knowledge of various distribution types, which someone without statistical training might not possess. Additionally, physical limitations such as color blindness can impede accurate interpretation if visualizations are not designed with accessibility in mind. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

@@

To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct a series of  empirical studies. First, we examine the outputs of code cells in Jupyter notebooks. Next, we study Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks. Finally, we investigate the intersection of outputs and assertions by analysing  assertions that document decisions made from the outputs of code cells. In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt. To the best of our knowledge, we are the first to investigate how feedback derived from the output of code cells is used by ML practitioners when developing ML pipelines inside of Jupyter notebooks. We are also the first to investigate assertions written inside of notebooks and their intersection with the outputs of code cells.

The research questions along with the contributions of this study are presented below.

\begin{description}
  \item[RQ1.] \textbf{What are the characteristics of feebback mechanisms in ML Jupyter notebooks?}

  % TODO: the problem is that readers don't yet know that outputs are generated from last statement and print statements
  We mine 300K Jupyter notebooks written in Python from Github and Kaggle. We analyse 2.3M code cells and create a dataset of 22.5K assertions, 25K print statements and 28K last statements. We further perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks.

  % TODO: The dataset is available publicly on ...

  \item[RQ2.] \textbf{How is implicit feedback from the output of code cells used when writing ML code in Jupyter notebooks?}

  We conduct individual case-studies with XXX code cell outputs. During each case-study, we analyse the raw output along with the corresponding code that produced the output to understand its purpose. Additionally, we anayse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.

  \item[RQ3.] \textbf{How are assertions used to validate ML code written in Jupyter notebooks?}

  Similarly, we conduct case-studies with XXX assert statements. We analyse the assertion along with the surrounding code to understand its purpose.

  \item[RQ4.] \textbf{How are assertions used to validate visual outputs in Jupyter notebooks?}

  Finally, we conduct case-studies with XXX visualisation and assertion (VA) pairs.
\end{description}

% TODO revisit this paragraph
Our findings indicate that outputs from code cells are extensively utilised for making critical design and implementation decisions during the MLDL. While manual validation of these outputs is common, it is also prone to inconsistencies thus highlighting the need for automated validation mechanisms. Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds, thereby reducing reliance on manual checks and enhancing the robustness of ML pipelines. However, the study also notes the limited use of assertions, with only 9.7\% of the 3 million notebooks analyzed incorporating them, suggesting a significant gap in current testing practices within the ML community. The findings underscore the necessity for developing ML-specific testing tools and integrating assertions more deeply into the ML development workflow to ensure consistent and reliable validation processes.

% TODO update replication package

@@

This section introduces relevant prior concepts for this study.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{nb.pdf}
  \label{fig:nb}
  \caption{Example Jupyter notebook with code and markdown cells, adapted from~\cite[Figure~1]{pimentel2019large-scale}.}
\end{figure}

Literate programming is a paradigm introduced by Donald Knuth that emphasizes writing code that is understandable by humans. In his seminal work, Knuth advocates combining source code with documentation in a single, cohesive narrative~\cite{knuth1984literate}. Computational notebooks are inspired by this paradigm, allowing for the integration of executable code, scientific equations, visualisations, and explanatory text within a single document. This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows. Project Jupyter is a prominent example of this approach, offering a suite of open-source tools for interactive computing. Jupyter Notebooks, in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.

Figure~\ref{fig:nb} shows an example Jupyter notebook with markdown and code cells. The data comprised within Jupyter notebooks is stored using the JSON schema specified by \emph{nbformat}~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically. A notebook consists of \emph{cells} which can be of three types---``markdown'', ``code'' and ``raw''. All cells of a notebook are stored as a list under the top-level \lstinline[language={}]$cells$ key in the JSON representation. Within this list, each cell is represented as a JSON object. The contents of the cell are stored as a list of strings inside the \lstinline[language={}]$source$ key.

\subsection{Outputs in Jupyter Notebooks}

Only code cells in Jupyter notebooks can generate outputs. As shown in Figure~\ref{fig:nb}, outputs can come from three main sources in the source code---print statements, the last statement in the code cell, and statements that create a visualisation. The outputs produced by code cells are stored as JSON objects under the \lstinline[language={}]$outputs$ key in the corresponding JSON representation of the notebook. Text outputs (such as those produced by the last statement of a cell or print statements) are stored under the \lstinline[language={}]$text$ key. Visual outputs have a sightly different JSON schema and are stored under the \lstinline[language={}]$data$ key. The image itself is represented as a base64 encoded string which is stored inside the \lstinline[language={}]$data$ object, under the \lstinline[language={}]$image/png$ key.

@@

An Abstract Syntax Tree (AST) is a tree representation of the abstract syntactic structure of source code. Each node in the tree represents a construct occurring in the source code. ASTs are widely used in compilers and code analysis tools to understand and manipulate code programmatically~\cite{wikipedia2024abstract}. The Python \emph{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an AST. This module allows for the examination and modification of code structure, making it valuable for tasks such as code analysis, optimisation, and transformation. The ast module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation~\cite{pythonast}.

In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class. The \lstinline{Assert} node includes the \lstinline{test} attribute---which contains the condition being asserted---and the \lstinline{msg} attribute---which holds an optional message displayed if the assertion fails. Print statements are treated as a call to the built-in \lstinline{print} function. Function calls are represented in the AST by the \lstinline{Call} class. The \lstinline{Call} node includes the \lstinline{func} attribute representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.

\section{Methodology}

@@

% The predominant form of obtaining feedback, while developing ML pipelines in Jupyter notebooks, is through the output of code cells. These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualisations. Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code. This idea draws from prior literature in software engineering that connects test-driven development (TDD) as a form of developing software using feedback.

@@

We check the validity of the underlying JSON structure in each notebook using the nbformat tool\footnote{https://nbformat.readthedocs.io/en/latest/}. Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from white literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.

@@

\subsection{(RQ1) What are the characteristics of feedback mechanisms in ML Jupyter notebooks?}

Of the total 298K notebooks collected in this study, 25K (8.5\%) contained at least one assertion. From the 8.5\% notebooks, we collected 89.6K assertions. 4.8K assertions (5\%) used methods from external testing libraries. Majority of the assertions however (85K or 95\%) were written using the \lstinline{assert} statement provided by the built-in Python standard library. We removed duplicate assertions for the remainder of the analysis, thus resulting in 27.1K assertions.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{other-test-methods.pdf}
  \caption{10 most common methods used in assertions written using external testing libraries.}
  \label{fig:other-test-methods}
\end{figure}

Figure~\ref{fig:other-test-methods} presents the 10 most common methods used in the 5\% assertions written using external testing libraries. We manually tracked the source of the methods to the numpy, pandas, torch and unittest libraries. The most frequently used methods include \lstinline{assert_almost_equal} and \lstinline{assert_equal}, each with over 250 instances. Methods such as \lstinline{assert_almost_equal}, \lstinline{assert_array_almost_equal} and \lstinline{AssertAlmostEqual} accommodate the stochastic nature of ML, where results can vary slightly due to inherent randomness in data processing and model training. We also find methods such as \lstinline{assert_array_equal} and \lstinline{assert_frame_equal} that are specifically tailored for comparing high-dimensional data structures such as numpy arrays and pandas data frames.

\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-assert-test.pdf}
    \caption{10 most common tests used in \lstinline{assert} statements.}
    \label{fig:common-assert-test}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-compare-op.pdf}
    \caption{10 most common operators used in \lstinline{assert} statements that perform a comparison check.}
    \label{fig:common-compare-op}
  \end{minipage}
\end{figure}

% TODO: some form of background on AST representation of assert statements is missing

\begin{table}
  \centering
  \caption{Examples of the most common tests performed in assert statements.}
  \begin{tabular}{@{}m{0.3\textwidth} m{0.3\textwidth}@{}}
    \toprule
    \emph{\textbf{Comparison Check}}&
    \emph{\textbf{Function Call}}\\
    \midrule

    \lstinline[]$type(train_dataset[0]) == tuple$&
    \lstinline[]$(np.unique(y_test) == [-1, 1]).all()$\\

    \lstinline[]$torch.__version__ > '1.10.0'$&
    \lstinline[]$isinstance(transmat, pd.DataFrame)$\\

    \lstinline[]$len(x_dev) == len(y_dev)$&
    \lstinline[]$np.allclose(np.linalg.norm(wio, axis=0), 1)$\\
    \bottomrule
  \end{tabular}
  \label{tab:common-assert-test}
\end{table}

We focus the remainder of the analysis on the 95\% of the assertions written using the \lstinline{assert} statement. The most common test performed in the \lstinline{assert} statements include a comparison check and a function call as shown in Figure~\ref{fig:common-assert-test}. Table~\ref{tab:common-assert-test} provides a few examples of assertions for each of the tests above. In Figure~\ref{fig:common-compare-op} we see the most common operators used in assert statements that perform a comparison check. We observe that overwhelming majority of the asserts ensure that two values are equal to each other. The shape and size of data structures are frequently compared as observed in Figure~\ref{fig:common-compare-lhs} and Figure~\ref{fig:common-compare-rhs} which show the most common bi-grams that appear in the LHS and RHS of the comparison statements respectively.

\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-compare-lhs.pdf}
    \caption{Most common bi-grams that appear in the LHS of comparison statements.}
    \label{fig:common-compare-lhs}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-compare-rhs.pdf}
    \caption{Most common bi-grams that appear in the RHS of comparison statements.}
    \label{fig:common-compare-rhs}
  \end{minipage}
\end{figure}

Of the 95\% assert statements, 28.4K (31.7\%) have a failure message. Figure~\ref{fig:common-assert-msgs} shows the most frequent bi-grams that appear in these failure messages. Similar to the analysis presented above, we removed the duplicate messages prior to analysing the bi-grams resulting in 5.9K unique messages. The bi-grams indicate that the shape and data type of data structures are frequently verified.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{common-assert-msgs.pdf}
  \caption{Most common bi-grams in the failure messages of assert statements.}
  \label{fig:common-assert-msgs}
\end{figure}

\subsection{(RQ2) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}

@@

Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis. For instance, visualisations can efficiently determine if scaling, normalizing, or handling outliers is needed. During the exploratory data analysis (EDA) phase, visualisations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training. Additionally, descriptive statistics are used post-transformation to verify the effectiveness of steps like data normalization, ensuring data conforms to the expected format for further analysis.

\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualisations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48). These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the EDA phase.}

@@

Table~\ref{tab:linear-relation-check} illustrates two outputs found in this study for identifying linear relationships amongst features in the data. In output \emph{O6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}. Output \emph{O10} depicts the visualisation of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.

Linear machine learning models achieve optimal performance when the target variable can be expressed as a linear combination of the input features. However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training. Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\cite{shome2022data}.

\highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data. visualisations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features. These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}

@@

Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system. This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.

@@

While training multiple machine learning models for a single task is common, choosing the best model considers not only its performance but also its training speed. Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{shome2022data}.

@@

Table~\ref{tab:shape-check} shows print statements discovered in this study which check the shape of the data. \emph{P4} prints the number of rows or examples in the test set. \emph{P32} and \emph{P117} prints the number of rows and columns (or features) in the training and test data respectively.

@@

\subsection{(RQ3) How are assertions used to validate ML code written in Jupyter notebooks?}~\label{sec:result-rq2}

@@

Table~\ref{tab:assert-resource-check} persents the assertions encountered in this study, that perform resource checks encompassing pre-trained models, data paths, visualisations, and model training status. These checks help prevent runtime errors and ensure the proper execution of code.

@@

Resource check assertions play a vital role in maintaining the reliability and robustness of the machine learning workflows by ensuring the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualisations.

@@

\subsection{Assertions and Software Bugs}
 TODO
